---
title: "Test multiple models"
output:
  rmarkdown::html_vignette:
    mathjax: default
vignette: >
  %\VignetteIndexEntry{Test multiple models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(logmu)
```

`fit()` calibrates one model. It cannot tell you whether that model was worth
having, because a model with more parameters always fits the data it was
calibrated on better than a model with fewer. `compare_models()` fits several
candidates and ranks them on penalised log-likelihood, which is Akaike's
information criterion with the clutter removed. The reasoning behind it is in
[Suddenly AIC](B050-suddenly-aic.html); this article is about running it.

## A portfolio to work on

There is no data in the package, so here is some, generated from a mortality we
will pretend not to know. Deaths are placed by inverting the integrated hazard,
which for a Gompertz $\mu$ can be done in closed form, so this is a sample from
the model rather than an approximation to one.

```{r data}
set.seed(20260910)

lives <- 4000
age_at_start <- sample(55:89, lives, replace = TRUE)
male <- runif(lives) < 0.6

# The truth, which the candidates below have to find: a level, an age slope
# steeper than the reference table's, and a loading on males.
log_mu_at_start <- -12.9 + 0.12 * age_at_start + 0.4 * male

exposure <- 5
target <- -log(runif(lives))
death_after <- log1p(0.12 * target / exp(log_mu_at_start)) / 0.12
died <- death_after < exposure

portfolio <- exp_data(
  list(
    birth     = datey::datey(2015 - age_at_start),
    male      = male,
    pension   = round(exp(rnorm(lives, log(12000), 0.7))),
    E2R_start = datey::datey(rep(2015, lives)),
    E2R_end   = datey::datey(2015 + pmin(death_after, exposure)),
    E2R_died  = died
  ),
  exp_start = datey::datey(2015),
  exp_end   = datey::datey(2020)
)

sum(died)
```

## Four candidates

Every candidate is a `model()`: a reference mortality, plus the covariates whose
coefficients are to be estimated. A `model` holds nothing else. The data, the
population, the weight and the overdispersion are arguments of the comparison,
not properties of a candidate, and that is what makes candidates comparable
rather than a rule you have to remember.

The reference here is a Gompertz curve with its time trend switched off, so
anything the candidates find is theirs.

```{r candidates}
reference <- gompertz_mortality(slope_t = 0)

candidates <- list(
  level    = model(reference, covariates(level = 1)),
  sex      = model(reference, covariates(male = .i$male, female = !.i$male)),
  age      = model(reference, covariates(level = 1, age = .x - 75)),
  sex_age  = model(reference, covariates(male = .i$male, female = !.i$male,
                                         age = .x - 75))
)
```

The `sex` and `sex_age` candidates split the population into two indicators
that cannot both be one. Saying so with `disjoint()` instead of `covariates()`
lets the engine leave the off-diagonal integrals out altogether, which is worth
having once there are more than a handful of groups.

## The comparison

```{r compare}
comparison <- compare_models(portfolio, candidates,
                             overdispersion = 1)
comparison
```

Larger $L_\text{P}$ is better, so the table is ordered best first. The header
describes the data and the scale everything below it is quoted in.

The two likelihood columns have the best candidate's $L_\text{P}$ taken off them,
so the winner reads 0 and everyone else is negative. Both are shifted by the same
constant, which is why $L - L_\text{P} = p$ still holds row by row: the
arithmetic can be checked on the face of the table. `$table` keeps the unshifted
numbers.

Note the `p` column against the `k` beside it: the penalty is exactly the
parameter count. That is not a coincidence and it is not hard-wired -- the
penalty is
$p = Z^{-1}\operatorname{tr}(\mathbf{J}\mathbf{I}^{-1})$, and with no weight
$w^2 = w$ makes it collapse to $\dim(\beta)$. So here, and only here,
$L_\text{P}$ is exactly $-\text{AIC}/2$ and a shortfall of 1 is one parameter's
worth. Anything under about 1 is not worth acting on.

The winner is a `fit` like any other:

```{r best}
best <- best_fit(comparison)
coef(best)
sqrt(diag(vcov(best)))
```

The two sex coefficients differ by 0.399, against the 0.4 that went into the
data. The age term comes out at 0.0149 where 0.02 went in, about one standard
error short -- the level and the slope are estimated jointly and should be read
that way, not one at a time.

The fitted model is a `mortality`, so `best$mortality` can be handed straight to
`aev()`, or used as the `ref_mortality` of a further model.

## Weights change what the penalty means

Weight the log-likelihood by pension and the ranking still works, but two things
change.

First, $Z$ is no longer 1, so it has to be measured. It is measured on
`test_mortality`, which defaults to `default_mortality()`, and it has to be the
same for every candidate or the penalised log-likelihoods are on different
scales and the comparison means nothing. `compare_models()` measures it a single
time and hands the same value to every candidate, whatever the reference
mortalities of the candidates themselves happen to be.

Nominate a different test mortality if you have a reason to. Only its age slope
materially matters, and a shared choice is worth more than a good one, so the
default is usually the right answer. Where you already have the number, pass
`Z` instead and no measurement happens at all.

Second, $p$ stops being a count.

```{r weighted}
weighted <- compare_models(portfolio, candidates,
                           weight = .i$pension,
                           test_mortality = reference,
                           overdispersion = 1)
weighted$table[, c("k", "penalty", "penalised_log_likelihood")]
```

Mostly it stops being a count. `level` still costs exactly 1, and that is not
rounding: with a single constant covariate $\mathbf{J}\mathbf{I}^{-1}$ is the
scalar $\mathrm{E}w^2/\mathrm{E}w$, which is $Z$ itself, so $p = Z^{-1}Z = 1$
however the data is weighted. The others sit a little above their term counts,
three parameters costing 3.03.

The interesting change is in the spread. `sex_age` now leads `sex` by 0.2 where on
lives it led by 4.3: weighting concentrates the data, so there is far less
evidence for the age term once the log-likelihood follows the money. The ranking
survives. What you no longer get for free is the calibration -- a shortfall of 1
is no longer one parameter's worth, and
[Suddenly AIC](B050-suddenly-aic.html#just-weight-a-moment) sets out why.

## Overdispersion favours simpler models

Mortality data is overdispersed, always, and `overdispersion` has no default for
that reason. It is not a nuisance in this calculation: it changes the answer, and
in the right direction.

$L$ carries $\Omega^{-1}$ and the penalty does not, because the $\Omega^{-1}$ in
$\mathbf{I}$ and $\mathbf{J}$ cancels in $\mathbf{J}\mathbf{I}^{-1}$. So raising
$\Omega$ shrinks the advantage a richer model has while leaving what its extra
parameters cost untouched.

```{r overdispersion}
behind <- function(omega) {
  table <- compare_models(portfolio, candidates, overdispersion = omega)$table
  max(table$penalised_log_likelihood) - table["level", "penalised_log_likelihood"]
}

c(omega_1 = behind(1), omega_5 = behind(5), omega_25 = behind(25))
```

At $\Omega = 1$ the level-only model is 17 behind. At 5 it is under 2. At 25 its
shortfall is zero, meaning it has won outright -- at that much overdispersion the
evidence for a sex split is not there, and a single level is all the data
supports. This is the quasi-AIC, and it is why $\Omega$ deserves an argument
rather than a default.

## When a candidate cannot be fitted

`fit()` raises rather than return a number nobody should use. A comparison
cannot afford to: one hopeless candidate out of fifty must not take the other
forty-nine with it. So a candidate that fails is reported in place -- its row
carries the status and `NA`, it sorts last, and the diagnosis `fit()` would have
raised is kept.

```{r failure}
broken <- c(candidates,
            list(collinear = model(reference,
                                   covariates(level = 1, again = 1))))

comparison <- compare_models(portfolio, broken, overdispersion = 1)
comparison$failures[["collinear"]]
```

The warning names the failures as they happen, so a candidate cannot go missing
from a ranking quietly.
